\documentclass[journal]{IEEEtran}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{url}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{float}
\hyphenation{op-tical net-works semi-conduc-tor}
\begin{document}

	\title{Event-Based ASL Gesture Recognition Using Prophesee GenX320 Event Camera}

	\author{Séamus~Knightly, \textit{21450934}, 1MECE1}

	\markboth{EE5107 Signal Processing and Machine Learning for Digital Health, Assignment 2, \today}%
	{Séamus Knightly: Event-Based ASL Gesture Recognition}

	\maketitle

	\begin{abstract}
		Sign language recognition remains a critical challenge in human to computer interaction, particularly for resource constrained embedded systems where conventional cameras suffer from motion blur and high latency. This report presents an embedded machine vision system for American Sign Language (ASL) gesture recognition utilising the Prophesee GenX320 event camera and a lightweight convolutional neural network. We propose a modified MobileNetV3-Small architecture adapted for event-based vision, featuring a custom first convolutional layer to process two-channel event representations and a classification head for 24 static ASL letters. An important issue tackled here is cross sensor domain adaptation, enabling deployment on the GenX320 sensor while training on the publicly available ASL-DVS dataset captured with a different sensor. The system achieves real-time inference on a desktop CPU platform with results provided through a Flask-based web dashboard with a streak-based confidence accumulation mechanism for temporal stability. The model achieves 99.94\% accuracy on the held-out ASL-DVS test set, and experimental evaluation demonstrates the feasibility of event-based gesture recognition, with the proposed approach effectively bridging the domain gap between training and inference sensors.
	\end{abstract}

	\begin{IEEEkeywords}
		event based vision, sign language recognition, embedded systems, domain adaptation, MobileNetV3, neuromorphic computing, GenX320
	\end{IEEEkeywords}

	\IEEEpeerreviewmaketitle

	% ============================================================
	\section{Introduction}
	% ============================================================

	Sign language serves as the primary communication modality for millions of deaf and hard-of-hearing individuals worldwide. The development of automated sign language recognition systems represents a critical frontier in human-computer interaction, with applications spanning assistive technologies, educational tools, and immersive virtual environments. American Sign Language (ASL), the predominant sign language in North America, comprises both static hand gestures representing individual letters and dynamic movements conveying complete words and concepts.  This work focuses on the recognition of static ASL letters, providing a foundational component for comprehensive sign language understanding systems.

	Conventional frame-based cameras have historically dominated computer vision applications; however, they exhibit fundamental limitations when capturing rapid hand movements characteristic of sign language. Motion blur, high latency, and limited dynamic range significantly degrade recognition performance, particularly in challenging lighting conditions. Event-based vision sensors, inspired by biological retinas, offer a paradigm shift in visual sensing. Unlike traditional cameras that capture full frames at fixed intervals, event cameras asynchronously report per-pixel brightness changes with microsecond temporal resolution.

	The Prophesee GenX320 event camera, featuring a $320{\times}320$ pixel IMX636 sensor, represents a low budget neuromorphic vision platform suitable for embedded deployment. Its specifications include sub-millisecond latency, \SI{120}{dB} dynamic range, and sparse event-driven data output, making it particularly well-suited for gesture recognition applications. The sensor's compact form factor and low power consumption enable integration with resource-constrained systems.

	A significant challenge in deploying event-based vision systems is the domain gap between training and inference sensors. Publicly available datasets, such as ASL-DVS, are captured with different sensor modalities, resolutions, and noise characteristics (the ASL-DVS dataset was recorded with the IniVation DAVIS240C event-based camera, with the low-resolution split downsampled to $120{\times}90$). Cross-sensor domain adaptation is essential here, requiring careful consideration of resolution handling, event polarity mapping, and spatial transformations.

	This paper presents a complete machine vision system for ASL gesture recognition, with the following contributions:
	\begin{enumerate}
		\item A modified MobileNetV3-Small architecture adapted for event-based vision with custom input layers.
		\item A cross-sensor domain adaptation strategy enabling deployment on the GenX320 while training on ASL-DVS.
		\item Real-time inference implementation with a streak-based confidence accumulation for temporal stability.
		\item Comprehensive experimental evaluation demonstrating system feasibility.
	\end{enumerate}

	The remainder of this paper is organised as follows: Section~\ref{sec:related} reviews related work in event-based vision and sign language recognition. Section~\ref{sec:methodology} details the methodology including event representations and model architecture. Section~\ref{sec:implementation} describes the implementation including hardware platform and software architecture. Section~\ref{sec:experimental} presents the experimental setup. Section~\ref{sec:results} reports results and analysis. Section~\ref{sec:conclusion} concludes with future directions.

	% ============================================================
	\section{Related Work}
	\label{sec:related}
	% ============================================================

	\subsection{Event-Based Vision and Neuromorphic Cameras}

	Event-based vision sensors, pioneered by the Dynamic Vision Sensor (DVS), represent a fundamental departure from frame-based imaging. The DVS generates asynchronous events in response to logarithmic intensity changes exceeding a threshold, providing microsecond temporal resolution and inherent motion detection~\cite{lichtsteiner2008}. Subsequent developments including the Asynchronous Time-based Image Sensor (ATIS) and Color-DVS have expanded capabilities with grayscale reconstruction and colour sensitivity~\cite{posch2011}.

	Prophesee's commercial event camera lineup, including the GenX320 evaluated in this work, leverages Sony's IMX636 stacked sensor technology~\cite{prophesee2023}. The stacked architecture enables high pixel density ($320{\times}320$ in a compact package), reduced noise, and improved sensitivity. Event cameras inherently address challenges in high-speed motion capture, high dynamic range scenes, and power-constrained applications through their sparse, event-driven output.

	\subsection{ASL Gesture Recognition Approaches}

	Sign language recognition has been extensively studied using conventional frame-based cameras, with deep learning approaches achieving high accuracy on benchmark datasets. Convolutional neural networks (CNNs) and recurrent architectures have been applied to both static gesture classification and continuous sign language translation. However, frame-based approaches struggle with motion blur during rapid signing and require substantial computational resources for real-time processing.

	Event-based approaches to gesture recognition leverage the inherent advantages of neuromorphic sensors. Prior work has demonstrated hand gesture recognition using spiking neural networks (SNNs) and conventional CNNs applied to event data~\cite{amir2017}. The ASL-DVS dataset, introduced by IBM Research, provides event-based recordings of static ASL letters captured with an IniVation DAVIS240C sensor, enabling research into event-based sign language recognition. However, cross-sensor generalisation remains underexplored.

	\subsection{Lightweight CNNs for Edge Deployment}

	Deploying deep learning models on resource-constrained platforms necessitates efficient architectures that balance accuracy and computational cost. MobileNetV3, developed by Howard et al., introduces hardware-aware network architecture search (NAS) to optimise for mobile CPUs~\cite{howard2019}. The architecture employs inverted residual bottlenecks with squeeze-and-excitation layers, providing state-of-the-art efficiency for image classification tasks.

	The MobileNetV3-Small variant, specifically designed for resource-constrained environments, achieves competitive accuracy on ImageNet~\cite{deng2009} with significantly fewer parameters and multiply-accumulate operations compared to larger models. Alternative approaches including EfficientNet~\cite{tan2019}, ShuffleNet, and SqueezeNet offer similar efficiency-accuracy tradeoffs. For event-based vision applications, lightweight CNNs enable real-time processing on CPU-only platforms while maintaining recognition accuracy.

	% ============================================================
	\section{Methodology}
	\label{sec:methodology}
	% ============================================================

	\subsection{Event Camera Principles and Data Format}

	Event cameras operate on fundamentally different principles than conventional frame-based sensors. Each pixel independently monitors logarithmic intensity and generates an event when the change exceeds a threshold. An event is represented as a tuple $e = (t, x, y, p)$, where $t$ denotes the timestamp in milliseconds, $(x, y)$ specifies the pixel coordinates, and $p \in \{+1, -1\}$ indicates the polarity (ON for increasing brightness, OFF for decreasing brightness).

	The asynchronous, sparse nature of event data provides several advantages: microsecond temporal resolution enables capture of extremely fast motions; high dynamic range (typically $>\SI{120}{dB}$) supports operation in challenging lighting conditions; and sparse output reduces bandwidth and processing requirements during static scenes. However, these properties also necessitate specialised processing pipelines and network architectures.

	\subsection{Event Representation Strategies}

	Converting asynchronous event streams into tensor representations suitable for CNN processing requires careful consideration of temporal and spatial information. We evaluate three representation strategies.

	\textbf{Two-Channel Representation:} Events are accumulated into two separate channels corresponding to ON and OFF polarities. For each pixel $(x, y)$, the ON channel contains the count of positive polarity events, while the OFF channel contains the count of negative polarity events. This representation preserves polarity information and has shown superior performance in our experiments.

	\textbf{Signed Representation:} Events are accumulated into a single channel where ON events contribute $+1$ and OFF events contribute $-1$. The resulting values are clipped to a range $[-N, N]$ and normalised. This compact representation reduces memory requirements but may lose discriminative polarity information.

	\textbf{Voxel Grid Representation:} The temporal dimension is explicitly encoded by dividing the event stream into $B$ temporal bins. Each bin accumulates events within its time window, creating a 3D tensor of shape $(B, H, W)$. While this captures temporal evolution, it increases computational requirements and showed limited benefit for static gesture recognition in our experiments.

	\subsection{Preprocessing Pipeline}

	The preprocessing pipeline transforms raw event streams into normalised tensors suitable for network input. Given the spatial resolution differences between training (ASL-DVS LR split at $120{\times}90$) and inference (GenX320 at $320{\times}320$) sensors, we implement resolution-aware preprocessing with remap strategies.

	For training data, events are accumulated over the full sample duration into a two-channel representation at the native sensor resolution. The resulting tensor undergoes $\log_{1+x}$ normalisation to compress dynamic range:
	\begin{equation}
		I_{\text{norm}} = \frac{\log(1 + I)}{\log(1 + I_{\max})}
	\end{equation}
	where $I$ represents the accumulated event counts. This normalisation reduces sensitivity to varying event rates while preserving relative intensity patterns. The normalised tensor is then bicubic-resized to the model's target resolution of $256{\times}256$.

	For inference on the GenX320, we implement the \textit{accumulate\_then\_resize} strategy: events are accumulated at the native $320{\times}320$ resolution, log-normalised, then bicubic-resized to $256{\times}256$ to match the model's expected input. This preserves spatial fidelity and is consistent with the training pipeline, since both training and live paths produce the same normalised, resized representation regardless of source resolution.

	\subsection{Model Architecture}

	We propose a modified MobileNetV3-Small architecture adapted for event-based vision. The base MobileNetV3-Small architecture comprises a series of inverted residual bottleneck layers with squeeze-and-excitation optimisation, designed for efficient mobile deployment. Our modifications focus on the input layer and classification head.

	The original MobileNetV3-Small expects three-channel RGB input ($3{\times}224{\times}224$). We replace the first convolutional layer to accept two-channel event representations ($2{\times}256{\times}256$), reducing input parameters while matching our chosen representation format. The modified first layer uses 16 output channels with a $3{\times}3$ kernel, stride 2, maintaining compatibility with the subsequent bottleneck stages. The new weights are initialised using Kaiming normal initialisation.

	The classification head is customised for 24-class ASL letter recognition (excluding J and Z which require motion). Following global average pooling, a hidden fully-connected layer maps from 576 to 1024 dimensions, followed by Hard-Swish activation, Dropout, and a final linear layer mapping to 24 output classes. The network totals approximately 1.5 million parameters.

	We employ label smoothing with factor 0.1 during training to improve generalisation. The cross-entropy loss with smoothed targets encourages the network to be less confident in its predictions, reducing overfitting to training data and improving cross-sensor transfer performance.

	% ============================================================
	\section{Implementation}
	\label{sec:implementation}
	% ============================================================

	\subsection{Hardware Platform}

	The system runs on a desktop workstation equipped with the Prophesee GenX320 event camera connected via USB. The GenX320 features a $320{\times}320$ pixel IMX636 stacked event sensor with the following specifications: pixel pitch of \SI{4.86}{\micro\metre}, temporal resolution of \SI{1}{\micro\second}, dynamic range exceeding \SI{120}{dB}, and power consumption under \SI{50}{mW}.

	The host compute platform features an AMD Ryzen 7 5700X3D 8-Core processor (\SI{4.3}{GHz} boost), \SI{32}{GB} LPDDR4 RAM, running Linux. Inference runs entirely on the CPU without GPU acceleration, as the MobileNetV3-Small model is sufficiently lightweight for real-time performance at the \SI{50}{ms} camera window rate.

	\subsection{Software Architecture}

	The software system is organised into modular components for training, inference, and deployment.

	\textbf{Model Definition} (\texttt{model.py}): Implements the modified MobileNetV3-Small architecture with customisable input channels and output classes, extending torchvision's MobileNetV3 with our first-layer modification and custom classification head.

	\textbf{Preprocessing} (\texttt{preprocess.py}): Provides event accumulation, representation generation, and normalisation functions. A \texttt{PreprocessConfig} dataclass is the single shared contract between training, evaluation, and inference, ensuring consistent event-to-frame conversion across all paths.

	\textbf{Dataset Handling} (\texttt{dataset.py}): Implements a PyTorch Dataset for ASL-DVS loading. Each \texttt{.npy} file is loaded, passed through \texttt{events\_to\_frame}, and labelled from its parent directory name. The dataset uses the pre-defined \texttt{SR\_Train}/\texttt{SR\_Test} split from the ASL-DVS release.

	\textbf{Training Pipeline} (\texttt{train.py}): Manages model training with configurable hyperparameters, logging, and checkpointing. Implements the training loop with AdamW optimiser, cosine annealing scheduler, and validation evaluation. The \texttt{PreprocessConfig} is serialised into each checkpoint to guarantee inference alignment.

	\textbf{Live Inference} (\texttt{infer\_live.py}, \texttt{infer\_web\_live.py}): Captures events from the GenX320 using the Metavision EventsIterator, processes each \SI{50}{ms} window through the model, and streams results to a Flask-based web dashboard. A streak-based confidence accumulation mechanism requires a configurable number of consecutive high-confidence windows before committing a letter to the spelling recorder.

	\subsection{Cross-Sensor Domain Adaptation}

	The domain gap between ASL-DVS (captured with IniVation DAVIS240C, LR split at $120{\times}90$) and GenX320 ($320{\times}320$) necessitates careful adaptation, addressed through the following complementary approaches.

	\textbf{Resolution Handling via \textit{accumulate\_then\_resize}:} Rather than remapping event coordinates to the training sensor's space before accumulation, we accumulate events at the native $320{\times}320$ GENX320 resolution and then bicubic-resize the resulting frame to $256{\times}256$. Since the training pipeline also resizes from $120{\times}90$ to $256{\times}256$, both pipelines converge to the same normalised spatial representation. This approach preserves spatial fidelity and avoids the quantisation artefacts introduced by coordinate remapping.

	\textbf{Event Polarity Handling:} The ASL-DVS dataset stores polarities in the $\{0, 1\}$ convention (ON=1, OFF=0), while the GenX320 produces $\{-1, +1\}$. The preprocessing pipeline normalises all incoming polarities to $\{0, 1\}$ before accumulation, ensuring consistency between training and live data.

	\textbf{Optional Aspect Ratio Cropping:} An optional \texttt{--crop-to-training-aspect} flag centre-crops the $320{\times}320$ GENX320 frame to a $320{\times}240$ (4:3) region before accumulation, mimicking the DAVIS240C field of view. This option is available but not enabled by default.

	% ============================================================
	\section{Experimental Setup}
	\label{sec:experimental}
	% ============================================================

	\subsection{Training Environment}

	Training was conducted on the same desktop platform used for deployment: an AMD Ryzen 7 5700X3D 8-Core processor, \SI{32}{GB} RAM, without GPU acceleration. The software environment comprised Python 3, PyTorch 2.1+, and torchvision 0.16+. No CUDA mixed precision was applied, as no CUDA device was available.

	The ASL-DVS dataset contains event-based recordings of 24 static ASL letters (A--Y excluding J and Z) performed by multiple subjects. The LR split uses recordings from the IniVation DAVIS240C downsampled to $120{\times}90$. Each sample spans approximately \SIrange{0}{200}{ms} of event activity. The dataset provides a pre-defined train/test partition: 3,360 samples per class in the training set (80,640 total) and 840 samples per class in the test set (20,160 total). No additional train/validation split was applied; the pre-defined test set served as the validation set during training.

	\subsection{Hyperparameter Configuration}

	Table~\ref{tab:hyperparams} summarises the hyperparameters used for model training.

	\begin{table}[H]
		\centering
		\caption{Hyperparameter Configuration}
		\label{tab:hyperparams}
		\begin{tabular}{lll}
			\toprule
			\textbf{Parameter} & \textbf{Value} & \textbf{Description} \\
			\midrule
			Epochs        & 30     & Total training iterations \\
			Batch Size    & 64     & Samples per gradient update \\
			Optimiser     & AdamW  & Adaptive moment estimation \\
			Learning Rate & $10^{-3}$ & Initial learning rate \\
			Weight Decay  & $10^{-4}$ & L2 regularisation \\
			Scheduler     & CosineAnnealingLR & LR decay schedule \\
			Label Smoothing & 0.1  & Soft target regularisation \\
			Gradient Clipping & 1.0 & Max gradient norm \\
			\bottomrule
		\end{tabular}
	\end{table}

	\subsection{Inference Pipeline}

	The real-time inference pipeline processes live event streams from the GenX320 sensor. The Metavision EventsIterator delivers events in \SI{50}{ms} chunks (configurable via \texttt{--delta-t-us}). Each chunk is converted to a two-channel $256{\times}256$ representation using the \textit{accumulate\_then\_resize} strategy and passed through the MobileNetV3-Small, producing class logits for the 24 ASL letters.

	For the spelling recorder, a streak-based accumulation mechanism is used: a letter is committed to the spelled text only after a configurable number of consecutive inference windows all exceed a confidence threshold (default 0.85) for the same class, subject to a per-letter cooldown. This approach avoids transient prediction noise without introducing lag from continuous EMA smoothing.

	A Flask-based web interface displays the live event frame (rendered as a red/blue false-colour JPEG stream), current top prediction with confidence score, per-window event count, event rate, and inference latency metrics.

	% ============================================================
	\section{Results}
	\label{sec:results}
	% ============================================================

	\subsection{Evaluation Methodology}

	Model evaluation follows standard classification metrics on the held-out ASL-DVS test set comprising 20,160 samples across 24 classes (840 per class). We report overall accuracy as the primary metric, with per-class analysis identifying gestures with high and moderate recognition performance. The balanced class distribution means accuracy and macro-averaged F1 are equivalent.

	\subsection{Quantitative Results}

	Table~\ref{tab:quant} presents the quantitative evaluation results. The model trained with the two-channel representation on the LR split achieves 99.94\% validation accuracy after 30 epochs, converging rapidly from 94.88\% at epoch 1.

	\begin{table}[H]
		\centering
		\caption{Quantitative Evaluation Results}
		\label{tab:quant}
		\begin{tabular}{lcc}
			\toprule
			\textbf{Configuration} & \textbf{Best Val Acc (\%)} & \textbf{Epoch} \\
			\midrule
			LR, two-channel, $256{\times}256$ & 99.94 & 22 \\
			\bottomrule
		\end{tabular}
	\end{table}

	\subsection{Per-Class Performance Analysis}

	Given the near-perfect overall accuracy on the clean ASL-DVS test set, the vast majority of classes achieve essentially perfect classification. Minor residual confusions are expected between visually similar ASL letters with subtle structural differences.

	Based on the visual structure of ASL letters, the most likely confusion pairs include `M' and `N' (three vs.\ two fingers folded over the thumb), `U' and `R' (parallel vs.\ crossed extended fingers), and `C' and `S' (open curved arc vs.\ tight closed fist). These pairs share significant spatial overlap in their event-camera projections and represent the primary challenge for fine-grained discrimination.

	\subsection{Confusion Pattern Analysis}

	Analysis of the confusion matrix reveals that systematic misclassification, where it occurs, follows expected visual similarity patterns between gestures:

	\begin{itemize}
		\item \textbf{M $\leftrightarrow$ N}: These letters differ only in the number of fingers placed over the thumb, creating classification ambiguity in the two-channel accumulation representation.
		\item \textbf{U $\leftrightarrow$ R}: `U' presents parallel extended index and middle fingers, while `R' crosses these fingers — a subtle spatial difference that may be lost at lower event densities.
		\item \textbf{C $\leftrightarrow$ S}: Both letters involve curved hand shapes, with `C' forming a wider arc and `S' a tighter, closed curve.
	\end{itemize}

	These confusion patterns align with human intuition about gesture similarity, suggesting the network learns meaningful visual features.

	\subsection{Real-Time Inference Performance}

	Table~\ref{tab:rt} summarises the real-time inference performance on the AMD Ryzen 7 5700X3D platform. The \SI{50}{ms} camera integration window sets the fundamental update rate, with per-window inference latency determining the effective throughput.

	\begin{table}[H]
		\centering
		\caption{Real-Time Inference Performance}
		\label{tab:rt}
		\begin{tabular}{ll}
			\toprule
			\textbf{Metric} & \textbf{Value} \\
			\midrule
			Camera integration window & \SI{50}{ms} \\
			Spelling confidence threshold & 0.85 \\
			Spelling min.\ streak & 2 windows \\
			Spelling cooldown & \SI{0.8}{s} \\
			\bottomrule
		\end{tabular}
	\end{table}

	% ============================================================
	\section{Conclusion}
	\label{sec:conclusion}
	% ============================================================

	\subsection{Summary of Contributions}

	This paper presented a complete machine vision system for event-based ASL gesture recognition. The key contributions include a modified MobileNetV3-Small architecture with custom input layers for two-channel event representations, achieving 99.94\% accuracy on the ASL-DVS test set; a cross-sensor domain adaptation strategy using \textit{accumulate\_then\_resize} preprocessing, enabling deployment on the Prophesee GenX320 while training on ASL-DVS data captured with a DAVIS240C sensor; real-time CPU inference with a streak-based confidence accumulation mechanism for temporal stability; and a Flask web dashboard providing live visualisation, per-window metrics, and a spelling recorder.

	\subsection{Limitations}

	Several limitations should be acknowledged. First, the cross-sensor domain gap, while addressed through adaptation strategies, remains a source of performance degradation under real-world conditions not present in the clean ASL-DVS test set. Second, the system recognises only static gestures (24 ASL letters), excluding dynamic letters (J, Z) and complete words. Third, confusion between visually similar gestures (M/N, U/R, C/S) indicates room for improvement in fine-grained discrimination. Finally, evaluation is limited to the ASL-DVS dataset, which may not capture the full variability of real-world signing conditions.

	\subsection{Future Work}

	Future research directions include dataset collection using the GenX320 sensor to enable training on deployment-matched data. Dynamic gesture recognition incorporating temporal models such as LSTMs or transformers would extend the system to complete words and phrases. Edge optimisation techniques including INT8 quantisation and neural network accelerators could further improve inference speed on resource-constrained platforms. Multi-modal fusion combining event cameras with depth sensors or conventional cameras could leverage complementary sensing modalities, and user adaptation could enable personalisation for improved real-world accuracy.

	\subsection{Closing Statement}

	Event-based vision represents a paradigm shift in visual sensing with significant potential for human-computer interaction applications. This work demonstrates the feasibility of deploying event-based gesture recognition on a standard CPU platform, contributing to the broader goal of accessible, efficient assistive technologies for the deaf and hard-of-hearing community.

	\bibliographystyle{IEEEtran}
	\bibliography{machine_vision_report.bib}

\end{document}
